{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-19 thesis/thesis.tex
daves final revisions to acks and concl, change month, funding statemence

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Practical Concurrency Testing} \\
\normalsize \vspace{1em}
\begin{tabular}{c}
{\em or: How I Learned to Stop Worrying and Love the Exponential Explosion}
\end{tabular}}
\author{Ben Blum}
\date{October 2018}
\Year{2018}
\trnumber{}

@@

\support{}
\disclaimer{}

@@

Professor Garth Gibson is largely responsible for shaping me into the researcher I am today.
%Thanks to Garth's guidance,
%this thesis has grown into something that past-me,
%first starting research with the simple vision ``make a better autograder for 410'',
%could not have then imagined.
Despite having primary research interests in other fields,
Garth enthusiastically took on the project,
and always pushed me to explore new problem domains,
to design scientifically thorough experiments,
to explain difficult concepts approachably,
to appreciate related work as charitably as possible,
and to seek guidance from industry people with relevant experience,
despite my often stubborn refusal to leave my own little comfort zone.
So, even when it took me sending 10+ email reminders to get your attention,
truly,
thanks for everything, Garth.

@@

{Many have collaborated with me directly on the research itself,
whom I thank here in approximate chapter order.
Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first other users
%besides myself
back in its prototype days.
Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in \Cref{chap:landslide}.
Professor Brandon Lucia provided valuable revisions to \Cref{chap:quicksand}'s content
that ultimately led to its publication at OOPSLA 2016.
Michael J. Sullivan double-checked the proofs therein.
Professor Timothy Zhu from Penn State University
and Professor Haryadi Gunawi and TA Kevin Zhao from the University of Chicago
graciously allowed me to use their students as guinea pigs
and helped to improve Landslide's
%stability and
robustness for use beyond CMU's walls,
greatly enriching \Cref{chap:education}'s evaluation.
David Blum and Skye Toor provided
invaluable advice towards the statistical analysis
used therein.
%that helped me evaluate the student
%grade distributions therein.
Mario Dehesa-Azuara generously contributed the transactional
% justify last line bc of manual page break
\unskip\parfillskip 0pt \par}

@@

\noindent
memory data structures
that make up about half of \Cref{chap:tm}'s evaluation suite.
%
Carlo Angiuli, Evan Cavallo, Ziv Scully, Jim McCann,
Matthew Maurer, Stefan Muller, Guillaume Didier,
Sol Boucher, and Michael J. Sullivan
helped me rehearse and polish the snake fight talk.

@@

I thank the Parallel Data Lab, my research group at CMU,
for supporting me at the PDL retreat every fall,
pushing me to refine my presentation skills and to network with strangers (now familiar faces) from industry.
%I also thank those %industry
I also thank the industry attendees
%from each of the PDL Consortium member companies
for their interest %, even in more theoretical work,
%and for the enlightening conversations resulting therefrom.
and enlightening conversations.
Joan Digney and Karen Lindenfelser
make a lot of gears turn behind the scenes at PDL,
and Deb Cavlovich and Catherine Copetas
do likewise for CSD as a whole.
%Thanks to the anonymous OOPSLA reviewer who went the extra mile to read my supplementary proofs
%and urged me to improve the formalism to be more foundational
%(the lack of any stronger criticism a silent vote of confidence for their overall correctness).
Thanks to Nicholas D. Matsakis for the mentorship %and collaboration
during my two summers as a Rust research intern at Mozilla,
and to the rest of the Rust community as well for their %extremely important
work
bringing better concurrent programs into the world. %from a different angle.
And to everyone at large who
%outside my %immediate
%research circle who
%whether at the retreat or otherwise,
insisted how cool it was that my research helps students:
thank you for helping me keep my head up during darker times.

%I am grateful to
The ThursDz Council,
my group of now-former grad student, Shadyside-bound (in spirit if not in person) restaurant-hoppers,
surrounded me with both friendship and mentorship over the years.
In particular,
%in alphabetical for lack of a better order,
Chris Martens,
Jason Reed,
Jim McCann,
Rob Simmons,
Tom Murphy 7,
and
William Lovas:
you each have been role models to me
perhaps more than you know.

@@

My parents, Eve and David Blum, and my grandparents, Margie Granach and Elsie Blum,
provided bottomless emotional support throughout my education.
I appreciate beyond words for the privileged life they provided
that allowed me the opportunity to pursue higher education.

Looking back on my 11 years at CMU,
I am especially grateful for the friendships made early on %in my undergrad years
that continued well past first graduation.
From those who stayed in Pittsburgh %with me
to those who %moved away but
kept in touch,
from my regular tabletop, board, and/or dance gaming partners
to those who %inexhaustibly
welcomed me with company and
a couch to sleep on during travel,
I thank
%-- and in any order other than alphabetical here would be far too cruel --
8 Gianfortoni,
Adam Blank,
Alan Vangpat,
Alex Yuschik,
Anand Subramanian,
Andrew Krieger,
Car Bauer,
Carolyn Sawyer,
Elizabeth Kemp,
Elly Fong-Jones,
Emily Forney,
Emily Leathers,
Emma Cating-Subramanian,
Eric Faust,
Gabe Routh,
Greg Hanneman,
Hazel Vird\'{o},
Jack Ferris,
Jake Lengyel,
Jason Deng,
Josh Keller,
Joshua Wise,
Josiah Boning,
Julia Tuttle,
Kartik Subramanian,
Kellie ``K2'' Medlin,
Laura Abbott,
Maija Mednieks,
Margaret Meyerhofer,
Michael A\texttt{rntz}enius,
Michael J. ``Sully'' Sullivan,
Miriam Zimmerman,
Naomi Saphra,
Nathaniel Wesley Filardo,
Rhett Lauffenburger,
Richard Ha,
Rory Glenn,
Ryan Pearl,
and
Todd Eisenberger
for their wonderful friendship.
I am equally grateful to the new friends I made during my graduate years,
whether through board/card/dance gaming, tea drinking, %and/or
movie watching,
and/or as more permanent Pittsburgh residents that I am extremely torn up about moving away from:
%leaving behind (for now at least):
Alexandra Lee Falk,
Alexis Dyer,
Barbara Jensen,
Brendan McShane,
Brian Saghy,
Carlo Angiuli,
Cassie Orr,
Dan Guzek,
Danny Gratzer,
David Renshaw,
Evan Cavallo,
Gabriele Farina,
Grant Wu,
Jenny Lin,
Karl Schulz,
Kevin Nguyen,
Kristy Gardner,
Lea Albaugh,
Marlena Abraham,
Matt Stanec,
Matthew Maurer,
Sarah R. Allen,
Sol Boucher,
Stefan Muller,
Zachary Waldman,
Ziv Scully,
and the rest of ThursDz already named previously.
And to anyone who may read this list
and feel even slightly dejected that their name was left out:
I owe you a nice dinner. Get in touch.

@@

\begin{itemize}
	\item In \Cref{chap:quicksand},
		the soundness of data-race-driven Iterative Deepening
		combined with Quicksand's reliance on state space estimation
		to effectively balance formal verification with heuristic bug-finding speed
		better than any prior work.
		% 未来は、
		Even as modern programming languages such as Rust \cite{rust-book}
		threaten to obsolete the data-race preemption point proof
		by statically forbidding data races to begin with,
		Iterative Deepening's general approach remains relevant alongside ICB \cite{chess-icb},
		each providing a different tradeoff between bug-finding and completion time,
		and future model checkers should offer a mixture of different kinds of partial verification results
		depending on the user's wishes.
	\item In \Cref{chap:education},
		the behind-the-scenes yield-loop blocking algorithm
		(itself admittedly heuristic)
		allowed Landslide to automatically instrument all student implementations
		with very few false positives,
		and exposing partial results such as data race reports
		was shown to sometimes help students reach deeper understanding.
		% 未来は、
		Although high-level undergraduate operating systems courses represent
		a very small fraction of all the world's programmers,
		let alone all those in higher education,
		Landslide's adventures in 15-410 argue that computer science classes
		should more intentionally teach principled debugging techniques and tools
		rather than let students struggle to reach correctness independently. %on their own.
		Also, any future debugging tool should heed my students' survey feedback
		and take friendly user interface design more seriously than I did with Landslide.
	\item In \Cref{chap:tm},
		the equivalence of HTM's interface to failure injection and global locks
		allowed Landslide to avoid both performance overhead and state space explosion when simulating aborts,
		and the heuristic use of abort code reduction and abstraction reduction
		allowed verifying algorithms and data structures up to thread/iteration counts previously infeasible.
		% retry set reduction, obv; it also applies to failure injection as well
		% even when synch is complicated, defining abstraction boundaries and verifying your shit separately
		% is ,,good
		% - even if you have to be careful that you're "proving the right thing"
		% 未来は、
		Emulating complex execution semantics with simpler, already-understood primitives,
		and the retry set reduction,
		both allowing substantial yet sound state space reduction,
		are the obvious take-aways for any future model checker wishing to tackle HTM.
		I expect the latter should easily apply to other opportunities for failure-injection beyond transactions as well,
		such as memory allocation, filesystems, and distributed systems.
		More in general,
		the verification section demonstrated the value of compartmentalizing one's verification efforts,
		checking modules separately against a common, well-understood interface
		to reach yet further up the exponential cliff.
\end{itemize}
}

>>>>

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Practical Concurrency Testing} \\
\normalsize \vspace{1em}
\begin{tabular}{c}
{\em or: How I Learned to Stop Worrying and Love the Exponential Explosion}
\end{tabular}}
\author{Ben Blum}
\date{December 2018}
\Year{2018}
\trnumber{}

@@

\support{This research was sponsored by the U.S. Army Research Office under grant
number W911NF0910273 and by Intel ISTC-CC.
The views and conclusions contained in this document are those of the author
and should not be interpreted as representing the official policies,
either expressed or implied, of any sponsoring institution, the U.S. government, or any other entity.}
\disclaimer{}

@@

Professor Garth Gibson is largely responsible for shaping me into the researcher I am today.
%Thanks to Garth's guidance,
%this thesis has grown into something that past-me,
%first starting research with the simple vision ``make a better autograder for 410'',
%could not have then imagined.
Despite having primary research interests in other fields,
Garth enthusiastically took on the project,
and always pushed me to explore new problem domains,
to design scientifically thorough experiments,
to explain difficult concepts approachably,
to appreciate related work as charitably as possible,
and to seek guidance from industry people with relevant experience,
despite my often stubborn refusal to leave my own little comfort zone.
So, despite the times I needed to send 10+ email reminders to get your attention:
truly,
thanks for everything, Garth.

@@

{Many have collaborated with me directly on the research itself,
whom I thank here in approximate chapter order.
Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first other users
%besides myself
back in its prototype days.
Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in \Cref{chap:landslide}.
Professor Brandon Lucia provided valuable revisions to \Cref{chap:quicksand}'s content
	that ultimately led to its publication at \mbox{OOPSLA}~2016.
Michael J. Sullivan double-checked the proofs therein.
Professor Timothy Zhu from Penn State University
and Professor Haryadi Gunawi and TA Kevin Zhao from the University of Chicago
graciously allowed me to use their students as guinea pigs
and helped to improve Landslide's
%stability and
robustness for use beyond CMU's walls,
greatly enriching \Cref{chap:education}'s evaluation.
David Blum and Skye Toor provided
invaluable advice towards the statistical analysis
used therein.
%that helped me evaluate the student
%grade distributions therein.
Mario Dehesa-Azuara generously contributed the transactional
% justify last line bc of manual page break
\unskip\parfillskip 0pt \par}

@@

\noindent
memory data structures
that make up about half of \Cref{chap:tm}'s evaluation suite.
%
Carlo Angiuli, Evan Cavallo, Ziv Scully, Jim McCann,
Matthew Maurer, Stefan Muller, Guillaume Didier,
Sol Boucher, and Michael J. Sullivan
helped me rehearse and polish the snake-fight talk.

@@

I thank the Parallel Data Lab, my research group at CMU,
for supporting me at the PDL retreat every fall,
pushing me to refine my presentation skills and to network with strangers (now familiar faces) from industry.
%I also thank those %industry
I also thank the industry attendees
%from each of the PDL Consortium member companies
for their interest %, even in more theoretical work,
%and for the enlightening conversations resulting therefrom.
and enlightening conversations.
Joan Digney and Karen Lindenfelser
make a lot of gears turn behind the scenes at PDL,
and Deb Cavlovich and Catherine Copetas
do likewise for CSD as a whole.
%Thanks to the anonymous OOPSLA reviewer who went the extra mile to read my supplementary proofs
%and urged me to improve the formalism to be more foundational
%(the lack of any stronger criticism a silent vote of confidence for their overall correctness).
Thanks to Nicholas D. Matsakis for the mentorship %and collaboration
during my two summers as a Rust research intern at Mozilla,
and to the rest of the Rust community as well for
%their extremely important work
bringing better concurrent programs into the world. %from a different angle.
And to everyone at large who
%outside my %immediate
%research circle who
%whether at the retreat or otherwise,
insisted how cool it was that my research helps students:
thank you for helping me keep my head up during darker times.

%I am grateful to
The ThursDz Council,
my group of now-former grad student, Shadyside-bound (in spirit if not in person) restaurant-hoppers,
surrounded me with both friendship and mentorship over the years.
In particular,
%in alphabetical for lack of a better order,
Chris Martens,
Jason Reed,
Jim McCann,
Rob Simmons,
Tom 7,
and
William Lovas:
you each have been role models to me
perhaps more than you know.

@@

My parents, Eve and David Blum, and my grandparents, Margie Granach and Elsie Blum,
provided bottomless emotional support throughout my career as a student. %education.
I appreciate beyond words for the privileged life they provided
that allowed me the opportunity to pursue higher education.

Looking back on my 11 years at CMU,
I am especially grateful for the friendships made early on %in my undergrad years
that continued well past first graduation.
From those who stayed in Pittsburgh %with me
to those who moved away but % can comment 'moved away but'
kept in touch,
from my regular tabletop, board, and/or dance gaming partners
to those who %inexhaustibly
welcomed me with company and
a couch to sleep on during travel,
I thank
%-- and in any order other than alphabetical here would be far too cruel --
8 Gianfortoni,
Adam Blank,
Alan Vangpat,
Alex Yuschik,
Anand Subramanian,
Andrew Krieger,
Car Bauer,
Carolyn Sawyer,
Elizabeth Kemp,
Elly Fong-Jones,
Emily Forney,
Emily Leathers,
Emma Cating-Subramanian,
Eric Faust,
Gabe Routh,
Greg Hanneman,
Hazel Vird\'{o},
Jack Ferris,
Jake Lengyel,
Jason Deng,
Josh Keller,
Joshua Wise,
Josiah Boning,
Julia Tuttle,
Kartik Subramanian,
Kellie ``K2'' Medlin,
Laura Abbott,
Maija Mednieks,
Margaret Meyerhofer,
Michael A\texttt{rntz}enius,
Michael J. ``Sully'' Sullivan,
Miriam Zimmerman,
Naomi Saphra,
Nathaniel Wesley Filardo,
Rhett Lauffenburger,
Richard Ha,
Rory Glenn,
Ryan Pearl,
and
Todd Eisenberger
for their wonderful friendship.
I am equally grateful to the new friends I made during my graduate years,
whether through board/card/dance gaming, tea drinking, %and/or
movie watching,
and/or as more permanent Pittsburgh residents that I am extremely torn up about moving away from:
%leaving behind (for now at least):
Alexandra Lee Falk,
Alexis Dyer,
Barbara Jensen,
Brendan McShane,
Brian Saghy,
Carlo Angiuli,
Cassie Orr,
Dan Guzek,
Danny Gratzer,
David Renshaw,
Evan Cavallo,
Gabriele Farina,
Grant Wu,
Jenny Lin,
Karl Schulz,
Kevin Nguyen,
Kristy Gardner,
Lea Albaugh,
Marlena Abraham,
Matt Stanec,
Matthew Maurer,
Sarah R. Allen,
Sol Boucher,
Stefan Muller,
Zachary Waldman,
Ziv Scully,
and the rest of ThursDz already named previously.
And to anyone who may read this list
and feel even slightly dejected that their name was left out:
I owe you a nice dinner. Get in touch.

@@

\begin{itemize}
	\item In \Cref{chap:quicksand},
		the soundness of data-race-driven Iterative Deepening
		combined with Quicksand's reliance on state space estimation
		effectively balanced formal verification with heuristic bug-finding speed
		better than any prior work.
		% 未来は、
		Even as modern programming languages such as Rust \cite{rust-book}
		threaten to obsolete the data-race preemption point proof
		by statically forbidding data races to begin with,
		Iterative Deepening's general approach remains relevant alongside ICB \cite{chess-icb},
		each providing a different tradeoff between bug-finding and completion time,
		and future model checkers should offer a mixture of different kinds of partial verification results
		depending on the user's wishes.
	\item In \Cref{chap:education},
		Landslide's behind-the-scenes yield-loop blocking algorithm
		%(itself admittedly heuristic)
		allowed it to automatically instrument all student implementations
		with no false positives
		(what false positives did exist were minor unrelated Landslide bugs,
		all fixed quickly whenever reported),
		and exposing partial results such as data race reports
		was shown to sometimes help students reach deeper understanding.
		% 未来は、
		Although high-level undergraduate operating systems courses represent
		a very small fraction of all programmers in higher education,
		let alone all those in the world,
		Landslide's adventures in 15-410 argue that computer science classes
		should more intentionally teach principled debugging techniques and tools
		rather than let students struggle to reach correctness independently. %on their own.
		Also, any future debugging tool should heed my students' survey feedback
		and take friendly user interface design more seriously than I did with Landslide.
	\item In \Cref{chap:tm},
		the equivalence of HTM's interface to failure injection and global locks
		allowed Landslide to avoid both performance overhead and state space explosion when simulating aborts,
		and heuristic use of abort code reduction and abstraction reduction
		allowed verifying algorithms and data structures up to thread/iteration counts previously infeasible.
		% retry set reduction, obv; it also applies to failure injection as well
		% even when synch is complicated, defining abstraction boundaries and verifying your shit separately
		% is ,,good
		% - even if you have to be careful that you're "proving the right thing"
		% 未来は、
		Emulating complex execution semantics with simpler, already-understood primitives,
		and the retry set reduction,
		both allowing substantial yet sound state space reduction,
		are the obvious take-aways for any future model checker wishing to tackle HTM.
		I expect the latter should easily apply to other opportunities for failure-injection beyond transactions as well,
		such as memory allocation, filesystems, and distributed systems.
		More in general,
		the verification section demonstrated the value of compartmentalizing one's verification efforts,
		checking modules separately against a common, well-understood interface
		to reach yet further up the exponential cliff.
\end{itemize}
}
